Specific examples

### 2023-03-27 inst/gitbook/_book/dads_manual.tex
manual big update

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere

@@

\documentclass[]{book}

@@

\usepackage{fixltx2e} % provides \textsubscript

@@

  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex

@@

  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}

@@

% use upquote if available, for straight quotes in verbatim environments

@@

% use microtype if available
\IfFileExists{microtype.sty}{%

@@

@@

\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}

@@

            pdftitle={dads manual},
            pdfauthor={Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}

@@

@@

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{dads manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
\date{2020-12-04}

@@

\texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots{}) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Checking if devtools is already installed}
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\CommentTok{## Installing the latest version directly from GitHub}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{figure}
\centering
\includegraphics{dads_structure.pdf}
\caption{Schematised summary of the \texttt{dads} package architecture}
\end{figure}

@@

\hypertarget{getting-started}{%
\chapter{Getting started}\label{getting-started}}

@@

\begin{itemize}
\tightlist
\item
  \texttt{"max.taxa"\ \ \ =\ n} stop when \texttt{n} taxa are generated;
\item
  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e. ``living'' taxa);
\item
  \texttt{"max.time"\ \ \ =\ n} stop when the simulated tree is \texttt{n} units of age old (these units are arbitrary);
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Setting a stopping rule to reach a maximum of 20 taxa}
\NormalTok{my_stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Running the birth death simulation}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## The tree object}
\NormalTok{my_tree}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Plotting it}
\KeywordTok{plot}\NormalTok{(my_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## my birth death parameters}
\NormalTok{my_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
                  \DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that here it is not necessary to specify \texttt{extinction\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinciton\ =\ 0} to \texttt{extinction\ =\ 1/3}). However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
\end{quote}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Generating a birth death tree with extinctions:}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params, }\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
\CommentTok{## Visualising the new tree}
\KeywordTok{plot}\NormalTok{(my_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Creating the traits object}
\NormalTok{my_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Which process is in here?}
\NormalTok{my_trait}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What does it look like?}
\KeywordTok{plot}\NormalTok{(my_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Simulate disparity and diversity}
\NormalTok{my_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params,}
                \DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule,}
                \DataTypeTok{traits    =}\NormalTok{ my_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What's in there}
\NormalTok{my_data}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Plotting the disparity and diversity}
\KeywordTok{plot}\NormalTok{(my_data)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Extracting the tree (a "phylo" object)}
\NormalTok{the_generated_tree <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{tree}
\KeywordTok{class}\NormalTok{(the_generated_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Extracting the data (a "matrix")}
\NormalTok{the_generated_data <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{data}
\KeywordTok{class}\NormalTok{(the_generated_data)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Make a default modifiers}
\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
\CommentTok{## What's in it?}
\NormalTok{default_modifiers}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Setting the simulation parameters}
\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
\NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
\NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{default_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
\NormalTok{default_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).

For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
We can create one for example that makes species go extinct if their ancestor have a negative trait value.
For that we need to create a modifiers object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
First we are going to create the modification function.
This function must intake the argument \texttt{x} and, in our case, return a logical value: \texttt{TRUE} is for speciate and \texttt{FALSE} is for go extinct.
We want this \texttt{modify} function to always return \texttt{FALSE} (go extinct) since it will be triggered by \texttt{condition} (we will see that in a minute).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Going extinct}
\NormalTok{staying.alive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Now we want this function to only trigger when an ancestor has a negative trait value.
We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}s understanding of what you which to asses:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Triggering a modification only if the ancestor trait is negative}
\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Note that we use the function \texttt{all} here to evaluate all traits (if the data is multidimensional).
We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
If you are an advances \texttt{dads} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{dads} called\ldots{} \texttt{speciation}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Making a modifier for species to go extinct if}
\CommentTok{## their ancestor's trait value is (or are) negative}
\NormalTok{negatives_extinct <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}
                        \CommentTok{## The speciation function (default)}
                        \DataTypeTok{speciation =}\NormalTok{ speciation,}
                        \CommentTok{## What to modify}
                        \DataTypeTok{modify     =}\NormalTok{ staying.alive,}
                        \CommentTok{## When to modify it}
                        \DataTypeTok{condition  =}\NormalTok{ negative.ancestor)}
\CommentTok{## What's in it?}
\NormalTok{negatives_extinct}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{biased_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives_extinct)}
\NormalTok{biased_data}
\end{Highlighting}
\end{Shaded}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}

@@

\documentclass[
]{book}

@@

@@

  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex

@@

  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}

@@

% Use upquote if available, for straight quotes in verbatim environments

@@

\IfFileExists{microtype.sty}{% use microtype if available

@@

\@ifundefined{KOMAClassName}{% if non-KOMA class

@@

    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

@@

  pdftitle={Specific examples},
  pdfauthor={Thomas Guillerme (guillert@tcd.ie)},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs

@@

% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}

@@

% Set default figure placement to htbp
\def\fps@figure{htbp}

@@

\usepackage{booktabs}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{Specific examples}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
\date{2023-03-27}

@@

\texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

@@

\hypertarget{who-is-this-manual-for}{%
\section{Who is this manual for?}\label{who-is-this-manual-for}}

This manual explains how \texttt{dads} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.

The first section \texttt{Getting\ started} is aimed for people with a \texttt{R} beginner level:
* you know what a package is - and how to install it;
* you know what a function is - and what arguments are;
* you have already (vaguely) heard that they are different types of objects in R - like matrices (\texttt{"matrix"}) or trees (\texttt{"phylo"});
* and you have already looked for documentation and bugs online.

The sections after that are a bit more advanced and requires the following level:
* you have already created functions before;
* you know how to subset elements of a list - i.e.~you understand the \texttt{list\$element} syntax;
* you are aware that both \texttt{TRUE\ ==\ 1} and \texttt{FALSE\ ==\ 0} are \texttt{TRUE};
* you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between \texttt{as.numeric(1)} and \texttt{as.integer(1)};
* you owe part of your life to Stack Overflow.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Checking if devtools is already installed}
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\CommentTok{\#\# Installing the latest version directly from GitHub}
\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{figure}
\centering
\includegraphics{dads_structure.png}
\caption{Schematised summary of the \texttt{dads} package architecture}
\end{figure}

\hypertarget{bdalgorithm}{%
\chapter{The birth-death algorithm}\label{bdalgorithm}}

\begin{quote}
If you want to get your hands dirty, you can go straight to the \protect\hyperlink{gettingstarted}{getting start section}, this following section just describes the process in pseudo-code.
\end{quote}

The \texttt{dads} algorithm is based on a modular version of the birth-death process algorithm.
The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.

The algorithm used in \texttt{dads} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{-1}
\tightlist
\item
  \textbf{Starting the process}: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
\item
  \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct. In a standard birth-death process this is just done randomly, however in \texttt{dads} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
  Then go to step 2.
\item
  \textbf{Growing the tree} (\texttt{waiting}): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\]. In \texttt{dads}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.}
  If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
\item
  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{dads} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
  Then go to step 4.
\item
  \textbf{Speciating} (\texttt{speciating}): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
\end{enumerate}

\begin{verbatim}
if
    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
then
    do speciate
else
    go extinct
\end{verbatim}

Again, in \texttt{dads}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
Then go to step 5.
5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
6. The simulation has stopped because it reached the required amount of lineages or time.

@@

\hypertarget{gettingstarted}{%
\chapter{Getting started}\label{gettingstarted}}

@@

\begin{itemize}
\tightlist
\item
  \texttt{"max.taxa"\ \ \ =\ n} stop when \texttt{n} taxa are generated;
\item
  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e.~``living'' taxa);
\item
  \texttt{"max.time"\ \ \ =\ n} stop when the simulated tree is \texttt{n} units of age old (these units are arbitrary);
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Setting a stopping rule to reach a maximum of 20 taxa}
\NormalTok{my\_stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Running the birth death simulation}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# The tree object}
\NormalTok{my\_tree}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Plotting it}
\KeywordTok{plot}\NormalTok{(my\_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# my birth death parameters}
\NormalTok{my\_params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
                  \DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that here it is not necessary to specify \texttt{speciation\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinction\ =\ 0} to \texttt{extinction\ =\ 1/3}). However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
You can find more information about setting up more complex birth death parameters \protect\hyperlink{makebdparams}{in this section}.
\end{quote}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Generating a birth death tree with extinctions:}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\CommentTok{\#\# Visualising the new tree}
\KeywordTok{plot}\NormalTok{(my\_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating the traits object}
\NormalTok{my\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Which process is in here?}
\NormalTok{my\_trait}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# What does it look like?}
\KeywordTok{plot}\NormalTok{(my\_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
                \DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule,}
                \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# What\textquotesingle{}s in there}
\NormalTok{my\_data}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Plotting the disparity and diversity}
\KeywordTok{plot}\NormalTok{(my\_data)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Extracting the tree (a "phylo" object)}
\NormalTok{the\_generated\_tree \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{tree}
\KeywordTok{class}\NormalTok{(the\_generated\_tree)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Extracting the data (a "matrix")}
\NormalTok{the\_generated\_data \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{data}
\KeywordTok{class}\NormalTok{(the\_generated\_data)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Make a default modifiers}
\NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
\CommentTok{\#\# What\textquotesingle{}s in it?}
\NormalTok{default\_modifiers}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Setting the simulation parameters}
\NormalTok{extinction\_}\DecValTok{02}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
\NormalTok{living\_}\DecValTok{20}\NormalTok{     \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
\NormalTok{BM\_trait      \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
\NormalTok{default\_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

Note however that the printing information is now updated to state that you've added a modifier (even though it's a default one).

For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
We can create one for example that makes species go extinct if their ancestor have a negative trait value.
For that we need to create a modifiers object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.

Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met (say the lineage trait value is negative).
In other words, if the lineages trait value is negative, make sure that \emph{x} is equal to 1 (so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met).

First we need to create a condition function to trigger this (non) speciation event.
We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}' understanding of what you want to assess:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Triggering a modification only if the ancestor trait is negative}
\NormalTok{negative.ancestor \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{\textless{}}\StringTok{ }\DecValTok{0}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that we use the function \texttt{all} here to evaluate all traits: i.e.~if the data has more than one trait we trigger the modification only if all the trait values are negative.
\end{quote}

The we are going to create the modification function.
This function must intake the argument \texttt{x} and, in our case, returns the same value no matter what: \texttt{1} so that speciation never happens when the \texttt{negative.ancestor} function is triggered.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Always go extinct}
\NormalTok{going.extinct \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
If you are an advances \texttt{dads} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{dads} called\ldots{} \texttt{speciation}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Making a modifier for species to go extinct if}
\CommentTok{\#\# their ancestor\textquotesingle{}s trait value is (or are) negative}
\NormalTok{negatives\_extinct \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}
            \CommentTok{\#\# If the following condition is met...}
            \DataTypeTok{condition =}\NormalTok{ negative.ancestor, }\CommentTok{\# Does the lineage has a negative ancestor?}
            \CommentTok{\#\# Then apply the following modifier...}
            \DataTypeTok{modify =}\NormalTok{ going.extinct, }\CommentTok{\# The species goes extinct}
            \CommentTok{\#\# To the the speciation process.}
            \DataTypeTok{speciation =}\NormalTok{ speciation) }\CommentTok{\# Here the speciation() process is default}

\CommentTok{\#\# What\textquotesingle{}s in it?}
\NormalTok{negatives\_extinct}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives\_extinct)}
\NormalTok{biased\_data}
\end{Highlighting}
\end{Shaded}
